---
title: '4. Calling Functions in R'
author:  "CT1100 - J. Duggan"
output:
  beamer_presentation:
    theme: "CambridgeUS"
    colortheme: "beaver"
    fonttheme: "structurebold"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(ggplot2)
```

## Functions
- A function is a group of instructions that:
  + takes input, 
  + uses the input to compute other value, and 
  + returns a result (Matloff 2009). 
- Functions are a fundamental building block of R
- Users of R should adopt the habit of creating simple functions which will make their work more effective and also more trustworthy (Chambers 2008). 
- Two approaches:
  + Use R functions
  + Write your own functions

## R - A function "ecosystem"
- John Chambers: “To understand computations in R, two slogans are helpful:
  + Everything that exists is an object.
  + Everything that happens is a **function call.**"
  
```{r,echo=T}
x <- 1:6
sqrt(x)
min(x)
mean(x)
```

## Calling a function to create a plot (from package ggplot2)
```{r,echo=T,fig.width=6, fig.height=2.6}
library(ggplot2)
ggplot(data=mpg)+
  geom_point(mapping=aes(x=displ,y=cty))
```

## Adding information to the plot...
```{r,echo=T,fig.width=6, fig.height=2.6}
library(ggplot2)
ggplot(data=mpg,mapping=aes(x=displ,y=cty))+
  geom_point()+geom_smooth(method = "lm")
```


## Writing Functions
- **function** (*arguments*) *expression*
- arguments gives the arguments, separated by commas.
- Expression (body of the function) is any legal R expression, usually enclosed in { }
- Last evaluation is returned
- return() can also be used, but usually for exceptions.

```{r,echo=T}
f <- function(x)x^2 # this function squares a vector
f(1:3)
``` 

## Challenge 1.4
Write a function that takes in a vector and returns a vector with no duplicates. Make use of the R function duplicated().
```{r,echo=T}
x <- c(1, 2, 3, 4, 5, 1)
duplicated(x)
``` 

## Function Arguments
- It is useful to distinguish between  formal arguments and the actual arguments
  + Formal arguments are the property of the function
  + Actual arguments can vary each time the function is called.
- When calling functions, arguments can be specified by
  + Complete name
  + Partial name
  + Position
- Guidelines (Wickham 2015)
  + Use positional mapping for the first one or two arguments (most commonly used)
  + Avoid using positional mapping for less commonly used attributes
  + Named arguments should always come after unnamed arguments

## Function Arguments - Example

```{r,echo=T}
f1 <- function(arg1, arg2, arg3) arg1 * arg2  + arg3
f1(2, 3, 4) # positional
f1(2, arg3=4,3) # name for arg3
f1(arg3=4, arg2=3, 2) # name for arg2, arg3
``` 


## Functions are objects
- Functions are first class objects, so they can be passed to other functions
- Provides flexibility, and widely used in R
```{r,echo=T}
f1 <- function(f,v)f(v) # f is a function object

f1(min,c(2,4,6,7))
f1(max,c(2,4,6,7))
``` 

## Challenge 1.4
Write a function that takes in a vector and returns a vector with no duplicates. Make use of the R function duplicated().
```{r,echo=T}
x <- c(1, 2, 3, 4, 5, 1)
duplicated(x)
``` 



## Summary
- Functions are a fundamental building block of R
- Functions:
  + are declared using the function reserved word 
  + are objects
- Functions can access variables within the environment where they are created
- Functionals are functions that takes a function as an input and returns a vector as output (can be used as a looping structure)
- The apply family in R are functionals (**apply**, **sapply**, **lapply**)
- The package **purrr** is now being used instead of **apply**

```{r,echo=F,fig.width=2, fig.height=2}
# this is a hack to call plot but make the plot tiny
plot(1:1,axes=FALSE,xlab = "", ylab="",cex = .001)
```



