---
title: '6. Exploring Data with ggplot2 (1)'
author:  "CT1100 - J. Duggan"
output:
  beamer_presentation:
    theme: "CambridgeUS"
    colortheme: "beaver"
    fonttheme: "structurebold"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(ggplot2)
library(dplyr)
```

## Data Exploration

“Data exploration is the art of looking at your data, rapidly generating hypotheses, quickly testing them, then repeating again and again and again.” (Wickham and Grolemund 2017).

```{r, echo=F,fig.width=4, fig.height=1.8}
ggplot(data = mpg) +   # specify the source tibble
  geom_point(mapping=aes(x=displ,  # map x, y vars
                         y=hwy,
                         colour=class))
```

## Data visualisation with **ggplot2**

“The simple graph has brought more information to the data analyst’s mind that any other device.” – John Tukey 

```{r,echo=T}
d <- ggplot2::mpg  # get a copy of mpg
glimpse(d)  # show structure and some data
```

## Fuel Economy Data Set (ggplot2::mpg)

This dataset contains a subset of the fuel economy data that the EPA makes available on http://fueleconomy.gov. It contains only models which had a new release every year between 1999 and 2008 - this was used as a proxy for the popularity of the car.

|**manufacturer**| car manufacturer| **drv** | drive type |
|:------|:----------|:-----------|:-----------|
|**model**| model name| **cty** | city miles per gallon |
|**displ**| engine disp (l)| **hwy** | highway miles per gallon |
|**year**| year of make| **fl** | fuel type |
|**model**| model name| **cty** | city miles per gallon |
|**cyl**| number of cylinders| **class** | "type" of car |
|**trans**| type of transm.|  | |

## Exploring Data
Generate a first graph to help answer the following question

- Do cars with big engines use more fuel than cars with small engines
- What might the relationship between engine size and fuel efficiency look like?
  + Positive or negative?
  + Linear or non-linear?
- Variable (scatter plot)
  + **displ**, a car engine size in litres (x)
  + **hwy**, a car's fuel efficiency on highway - (y)
- ggplot2: layered approach
  + ggplot(data=*tibble_name*) + geom_point(mapping=aes(x=*col1*,y=*col2*))

## Plotting with ggplot2

```{r,echo=T,fig.width=5, fig.height=2.5}
ggplot(data = d) +   # specify the source tibble
  geom_point(mapping=aes(x=displ,  # map x, y vars
                         y=hwy))
```

## Interpreting the plot
- The plot shows a negative relationship between engine size (displ) and fuel efficiency (hwy)
- Cars with big engines use more fuel
- Does this confirm or refute your hypothesis about fuel efficiency and engine size?
```{r,echo=F,fig.width=4, fig.height=2.1}
ggplot(data = d,aes(x=displ, y=hwy)) +  
  geom_point() 
```

## Aesthetic Mappings
- A third variable can be added to a 2-D plot by mapping it to an aesthetic.
- An aesthetic is a visual property of the plot’s objects.
- An aesthetic’s level could be colour, size or shape


```{r,echo=T}
unique(d$class)
```

## In ggplot2 - Adding the third variable
```{r,echo=T,fig.width=5, fig.height=2.5}
ggplot(data=d)+
  geom_point(aes(x=displ,y=hwy,colour=class))
```

## Exploring Data Relationships
|Input (X)| Output (Y)| Hypothesis?|Reason |
|:------:|:----------|:----------|:----------|
|Displacement| City MPG| Negative? |Bigger cars,less efficient|
|Highway MPG| City MPG| Positive? |Should be closely related |
|Cylinders| Highways MPG| Negative? |More cylinders, less eff.|
|Cylinders| Displacement| Negative? |More cylinders, bigger eng.|

## x=displ, y=cty
```{r, echo=T,fig.width=5, fig.height=2.3}
ggplot(data = mpg) +   # specify the source tibble
  geom_point(mapping=aes(x=displ,  # map x, y vars
                         y=cty))
```

## x=hwy, y=cty
```{r, echo=T,fig.width=5, fig.height=2.3}
ggplot(data = mpg) +   # specify the source tibble
  geom_point(mapping=aes(x=hwy,  # map x, y vars
                         y=cty))
```

## x=cyl, y=cty
```{r, echo=T,fig.width=5, fig.height=2.3}
ggplot(data = mpg) +   # specify the source tibble
  geom_point(mapping=aes(x=cyl,  # map x, y vars
                         y=cty))
```

## x=cyl, y=displ
```{r, echo=T,fig.width=5, fig.height=2.3}
ggplot(data = mpg) +   # specify the source tibble
  geom_point(mapping=aes(x=cyl,  # map x, y vars
                         y=displ))
```

## Challenge 6.1
- Redraw the graphs, and colour by **car class**
- Vary the size of the point by using the number of cylinders

|Input (X)| Output (Y)| Hypothesis?|Reason |
|:------:|:----------|:----------|:----------|
|Displacement| City MPG| Negative? |Bigger cars,less efficient|
|Highway MPG| City MPG| Positive? |Should be closely related |
|Cylinders| Highways MPG| Negative? |More cylinders, less eff.|
|Cylinders| Displacement| Negative? |More cylinders, bigger eng.|

## Summary
- “The simple graph has brought more information to the data analyst’s mind that any other device.” – John Tukey ]
- “Data exploration is the art of looking at your data, rapidly generating hypotheses, quickly testing them, then repeating again and again and again.” (Wickham and Grolemund 2017).
- ggplot2 provides a layered approach to building charts

```{r,echo=F,fig.width=2, fig.height=2}
# this is a hack to call plot but make the plot tiny
plot(1:1,axes=FALSE,xlab = "", ylab="",cex = .001)
```



